---
title: "Batch effects, technical variables, and unwanted variation"
author: "Paulina Duda"
output: html_notebook
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/home/linux/PycharmProjects/pythonProject/CBS/Lab4&5")
```
#### Dependencies
```{r, results=FALSE, warning=FALSE, message=FALSE}
library("jackstraw")
library("SeuratData")
library("edge")
library("genefilter")
library("GEOquery")
library("sva")
library("Seurat")
library("tidyverse")
library("devtools")
library("patchwork")
library("RColorBrewer")
library("gplots")
library("data.table")
library("broom")
library("Biobase")
library("limma")
library("limma")
library("qvalue")
library("irlba")
library("Rtsne")
library("corpcor")
library("bladderbatch")
```
#### Load the `ExpressionSet` data
Make sure to apply log2 transformation and remove genes whose expression levels are below a threshold, 10:

```{r}
con = url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bottomly_eset.RData")
load(file=con)
close(con)
edata <- as.matrix(exprs(bottomly.eset))
dim(edata)
edata[1:5,1:5]
edata <- edata[rowMeans(edata) > 10, ]
edata <- log2(as.matrix(edata) + 1)
```

#### Bladder Cancer Gene Expression
```{r}
data(bladderdata)
pheno = pData(bladderEset)
edata = exprs(bladderEset)
dim(pheno)
dim(edata)
edata[1:5,1:10]
head(pheno)
```

#### Empirical Bayes
#### Using ComBat to clean a dataset
```{r}
batch = pheno$batch
combat_edata = ComBat(dat=edata, batch=pheno$batch, mod=model.matrix(~1, data=pheno), par.prior=TRUE, prior.plots=TRUE)
dim(combat_edata)
combat_edata[1:10,1:10]
```
Evaluate if the cleaned data from ComBat has no relation to batch effects:
```{r}
svd.out.combat <- svd(combat_edata)
svd.combat.plot <- data.table(svd.out.combat$v[,1:10], pheno)

ggplot(svd.combat.plot) + geom_point(aes(x=V1, y=V2, col=as.factor(batch)))
```
>Homework Problem 2: Make heatmaps, BEFORE and AFTER cleaning the data using ComBat, where columns are arranged according to the study design. You must sort the columns such that 5 batches are shown. Cluster the rows, but do not cluster the columns (samples) when drawing a heatmap.

BEFORE:
```{r}
my_palette <- colorRampPalette(c("blue", "white", "orange"))(n = 299)
# png("Duda_problem2.1.png",height=700,width=700)
heatmap.2(edata[,order(pheno$batch)],
          main = "Bladder Cancer Data Clustered", # heat map title
          notecol="black",      # change font color of cell labels to black
          density.info="none",  # turns off density plot inside color legend
          trace="none",         # turns off trace lines inside the heat map
          col=my_palette,       # use on color palette defined earlier
          dendrogram="none",     # only draw a row dendrogram
          scale="col",
          Rowv = TRUE,
          Colv = FALSE)
# dev.off()
```
AFTER:
```{r}
# png("Duda_problem2.2.png",height=700,width=700)
heatmap.2(combat_edata[,order(pheno$batch)],
          main = "Clustered and Cleaned by ComBat", # heat map title
          notecol="black",      # change font color of cell labels to black
          density.info="none",  # turns off density plot inside color legend
          trace="none",         # turns off trace lines inside the heat map
          col=my_palette,       # use on color palette defined earlier
          dendrogram="none",     # only draw a row dendrogram
          scale="col",
          Rowv = TRUE,
          Colv = FALSE)
# dev.off()
```
> Homework Problem 3: Make heatmaps of Pearson correlations statistics of samples.  First, compute the correlation statistics among columns. Second, create a heatmap using heatmap.2(). Make sure to create or add labels for samples (cancer vs. normal; batch numbers; others).
```{r}
# png("Duda_problem3.png",height=700,width=700)
heatmap.2(cor(edata, method = "spearman"),
          main = "Spearman correlations of samples", # heat map title
          notecol="black",      # change font color of cell labels to black
          density.info="none",  # turns off density plot inside color legend
          trace="none",         # turns off trace lines inside the heat map
          col=my_palette,       # use on color palette defined earlier
          dendrogram="none",    # only draw a row dendrogram
          scale="none",
          Rowv = FALSE,
          Colv = FALSE)
# dev.off()
```
>Homework Problem 4: Apply two different Linear Models to the Bottomly et al. data. First, using a conventional approach, create a linear model with a genetic strain (biological variable) and an experimental number (technical variable) on uncorrected gene expression data. Second, create a linear model with a genetic strain (biological variables) on corrected gene expression data from ComBat. Make a scatter plots of coefficients and a histogram of p-values as done in this notebook. Make sure that you are pulling out the correct coefficients, not any or all coefficients.

Unorrected gene expression data
```{r, message=FALSE}
pheno = pData(bottomly.eset)
edata = exprs(bottomly.eset)
edata <- edata[rowMeans(edata) > 10, ]
edata <- log2(as.matrix(edata) + 1)

mod = lm(t(edata) ~ as.factor(pheno$strain) + as.factor(pheno$experiment.number))
mod_tidy <- tidy(mod)
ggplot(mod_tidy %>% filter(term == 'as.factor(pheno$strain)DBA/2J')) +
    geom_histogram(aes(x=p.value), bins = 100, fill="darkorange") +
    ggtitle('LM Bottomly uncorrected, p-values histogram') +
    theme_light()
ggsave('duda_problem4.1.pdf')

ggplot(mod_tidy, aes(estimate, term)) +
     geom_point() +
     geom_vline(xintercept = 0) + ggtitle('LM Bottomly uncorrected, coefficients scatter plot') +
     theme_light()
ggsave('duda_problem4.2.pdf')
```
Corrected gene expression data from ComBat
```{r}
combat_edata = ComBat(dat=edata, batch=pheno$experiment.number,
               mod=model.matrix(~1, data=pheno), par.prior=TRUE, prior.plots=TRUE)
```
```{r, message=FALSE}

modcombat = lm(t(combat_edata) ~ as.factor(pheno$strain))
modcombat_tidy <- tidy(modcombat)
ggplot(modcombat_tidy %>% filter(term == 'as.factor(pheno$strain)DBA/2J')) +
    geom_histogram(aes(x=p.value), bins = 100, fill="darkorange") +
    ggtitle('LM Bottomly corrected, p-values histogram') +
    theme_bw()

ggsave('duda_problem4.3.pdf')

ggplot(modcombat_tidy, aes(estimate, term)) +
     geom_point() +
     geom_vline(xintercept = 0) +
     ggtitle('LM Bottomly corrected, coefficients scatter plot') +
     theme_bw()

ggsave('duda_problem4.4.pdf')
```

```{r, message=FALSE}
est_compare <- tibble(
  LinearModel = mod_tidy %>% filter(term == "as.factor(pheno$strain)DBA/2J") %>% select("p.value") %>% unlist,
  ComBat = modcombat_tidy %>% filter(term == "as.factor(pheno$strain)DBA/2J") %>% select("p.value") %>% unlist)

ggplot(est_compare, aes(x=LinearModel, y=ComBat)) +
     geom_point(col="darkgrey", alpha=.5, size=.5) +
     geom_abline(intercept=0, slope=1, col="darkred") +
     geom_smooth(method = "lm", se = TRUE) +
     ggtitle('LM coefficienties filtered vs unfiltered') +
     theme_bw()
ggsave('duda_problem4.5.pdf')

pvalues <- tibble(
  LinearModel = mod_tidy %>% filter(term == "as.factor(pheno$strain)DBA/2J") %>% select("p.value") %>% unlist,
  ComBat = modcombat_tidy %>% filter(term == "as.factor(pheno$strain)DBA/2J") %>% select("p.value") %>% unlist)

pvalues_gather = gather(pvalues)
ggplot(pvalues_gather, aes(x=value)) +
    geom_histogram() +
    facet_wrap(~key) +
    theme_bw()
ggsave('duda_problem4.6.pdf')
```
>Homework Problem 5: Apply ComBat and SVA to the Bottomly et al. data. Make a scatter plots of coefficients and a histogram of p-values, comparing results based on ComBat and SVA. Assume that the biological variables in Bottomly et al data is the genetic strains. Make sure that you are pulling out the correct coefficients/pvalues, not any or all of them.

Combat
```{r}
combat_edata = ComBat(dat=edata, batch=pheno$experiment.number,
               mod=model.matrix(~1, data=pheno), par.prior=TRUE, prior.plots=TRUE)
```
```{r, message=FALSE}
modcombat = lm(t(combat_edata) ~ as.factor(pheno$strain))
modcombat_tidy <- tidy(modcombat)
ggplot(modcombat_tidy %>% filter(term == 'as.factor(pheno$strain)DBA/2J')) +
    geom_histogram(aes(x=p.value), bins = 100, fill="darkorange") +
    ggtitle('LM Bottomly Combat-corrected, p-values histogram') +
    theme_bw()

ggsave('duda_problem5.1.pdf')

ggplot(modcombat_tidy, aes(estimate, term)) +
     geom_point() +
     geom_vline(xintercept = 0) +
     ggtitle('LM Bottomly Combat-corrected, coefficients scatter plot') +
     theme_bw()

ggsave('duda_problem5.2.pdf')
```

SVA
```{r, message=FALSE}
mod = model.matrix(~as.factor(strain), data=pheno)
mod0 = model.matrix(~1, data=pheno)
sva_output = sva(edata, mod, mod0, n.sv=num.sv(edata, mod, method="leek"))
modsva = lm(t(edata) ~ as.factor(pheno$strain) + sva_output$sv)
modsva_tidy <- tidy(modsva)
ggplot(modsva_tidy %>% filter(term == 'as.factor(pheno$strain)DBA/2J')) +
    geom_histogram(aes(x=p.value), bins = 100, fill="darkorange") +
    ggtitle('LM Bottomly SVA-corrected, p-values histogram') +
    theme_bw()

ggsave('duda_problem5.3.pdf')

ggplot(modcombat_tidy, aes(estimate, term)) +
     geom_point() +
     geom_vline(xintercept = 0) +
     ggtitle('LM Bottomly SVA-corrected, coefficients scatter plot') +
     theme_bw()

ggsave('duda_problem5.4.pdf')
```
Comparing
```{r, message=FALSE}
compare <- tibble(
  ComBat = modcombat_tidy %>% filter(term == "as.factor(pheno$strain)DBA/2J") %>% select("estimate") %>% unlist,
  SVA = modsva_tidy %>% filter(term == "as.factor(pheno$strain)DBA/2J") %>% select("estimate") %>% unlist)
pvalues <- tibble(
  ComBat = modcombat_tidy %>% filter(term == "as.factor(pheno$strain)DBA/2J") %>% select("p.value") %>% unlist,
  SVA = modsva_tidy %>% filter(term == "as.factor(pheno$strain)DBA/2J") %>% select("p.value") %>% unlist)

ggplot(compare, aes(x=ComBat, y=SVA)) +
     geom_point(col="darkgrey", alpha=.5, size=.5) +
     geom_abline(intercept=0, slope=1, col="darkred") +
     geom_smooth(method = "lm", se = TRUE) +
     ggtitle('LM coefficienties ComBat vs SVA') +
     theme_bw()

ggsave('duda_problem5.5.pdf')
```
```{r, message=FALSE}
pvalues_gather = gather(pvalues)

ggplot(pvalues_gather, aes(x=value)) +
    geom_histogram() +
    facet_wrap(~key) +
    theme_bw()

ggsave('duda_problem5.6.pdf')
```
