---
title: "Seurat"
author: "Paulina Duda"
output: html_notebook
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# setwd("/home/linux/PycharmProjects/pythonProject/CBS/Lab10")
```
#### Dependencies
```{r, results=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(Seurat)
library(patchwork)
library(Rtsne)
library(data.table)
library(ggplot2)
library(factoextra)
```
```{r, warning=FALSE}
pbmc.data <- Read10X(data.dir = "./filtered_gene_bc_matrices/hg19/")
pbmc <- CreateSeuratObject(counts = pbmc.data, project = "fresh68kpbmc", min.cells = 3, min.features = 200)
pbmc
```

#### Standard pre-processing workflow
#### QC and selecting cells for further analysis
```{r}
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
```
```{r}
VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```
```{r}
plot1 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```

#### Normalizing the data
```{r}
pbmc <- NormalizeData(pbmc, normalization.method = "LogNormalize", scale.factor = 10000)
pbmc <- NormalizeData(pbmc)
```

#### Identification of highly variable features (feature selection)
```{r}
pbmc <- FindVariableFeatures(pbmc, selection.method = "vst", nfeatures = 2000)

top10 <- head(VariableFeatures(pbmc), 10)

plot1 <- VariableFeaturePlot(pbmc)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2
```

#### Scaling the data
```{r}
all.genes <- rownames(pbmc)
pbmc <- ScaleData(pbmc, features = all.genes)
```

#### Perform linear dimensional reduction
```{r}
pbmc <- RunPCA(pbmc, features = VariableFeatures(object = pbmc))
print(pbmc[["pca"]], dims = 1:5, nfeatures = 5)
VizDimLoadings(pbmc, dims = 1:2, reduction = "pca")
DimPlot(pbmc, reduction = "pca")
DimHeatmap(pbmc, dims = 1, cells = 500, balanced = TRUE)
```

#### Determine the ‘dimensionality’ of the dataset
```{r}
pbmc <- JackStraw(pbmc, num.replicate = 100)
pbmc <- ScoreJackStraw(pbmc, dims = 1:20)
JackStrawPlot(pbmc, dims = 1:15)
ElbowPlot(pbmc)
```

#### Cluster the cells
```{r}
pbmc <- FindNeighbors(pbmc, dims = 1:10)
pbmc <- FindClusters(pbmc, resolution = 0.5)
head(Idents(pbmc), 5)
```

#### Run non-linear dimensional reduction (UMAP/tSNE)
```{r}
pbmc <- RunUMAP(pbmc, dims = 1:10)
DimPlot(pbmc, reduction = "umap")
```

#### Finding differentially expressed features (cluster biomarkers)
```{r}
cluster2.markers <- FindMarkers(pbmc, ident.1 = 2, min.pct = 0.25)
head(cluster2.markers, n = 5)
```
```{r}
cluster5.markers <- FindMarkers(pbmc, ident.1 = 5, ident.2 = c(0, 3), min.pct = 0.25)
head(cluster5.markers, n = 5)
```
```{r}
pbmc.markers <- FindAllMarkers(pbmc, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
pbmc.markers %>%
    group_by(cluster) %>%
    slice_max(n = 2, order_by = avg_log2FC)
```
```{r}
cluster0.markers <- FindMarkers(pbmc, ident.1 = 0, logfc.threshold = 0.25, test.use = "roc", only.pos = TRUE)
VlnPlot(pbmc, features = c("MS4A1", "CD79A"))
VlnPlot(pbmc, features = c("NKG7", "PF4"), slot = "counts", log = TRUE)
FeaturePlot(pbmc, features = c("MS4A1", "GNLY", "CD3E", "CD14", "FCER1A", "FCGR3A", "LYZ", "PPBP",
    "CD8A"))
```
```{r}
pbmc.markers %>%
    group_by(cluster) %>%
    top_n(n = 10, wt = avg_log2FC) -> top10
DoHeatmap(pbmc, features = top10$gene) + NoLegend()
```

#### Assigning cell type identity to clusters
```{r}
new.cluster.ids <- c("Naive CD4 T", "CD14+ Mono", "Memory CD4 T", "B", "CD8 T", "FCGR3A+ Mono",
    "NK", "DC", "Platelet")
names(new.cluster.ids) <- levels(pbmc)
pbmc <- RenameIdents(pbmc, new.cluster.ids)
DimPlot(pbmc, reduction = "umap", label = TRUE, pt.size = 0.5) + NoLegend()
```
## Homework
```{r}
pbmc.embedet <- Embeddings(object = pbmc, reduction = "pca")[,1:20]
k.mean <- kmeans(pbmc.embedet, 10)
k.mean.clusters <- k.mean$cluster
set.seed(1)
tsne_out <- Rtsne(pbmc.embedet,pca=F,perplexity=30)
tsne_out_pos = data.table(tsne_out$Y)
tsne_out_pos$cluster <- k.mean$cluster

ggplot(tsne_out_pos) +
        geom_point(aes(x=V1, y=V2, col = as.factor(cluster))) +
        labs(color = "cluster") +
        theme_bw()

```
```{r}
cluster.1 <- pbmc.embedet[k.mean.clusters == 1,]
cluster.1.kmean <- kmeans(cluster.1, 5)
clusters.cluster.1 <- cluster.1.kmean$cluster
cluster.1.tsne <- Rtsne(cluster.1, pca = F, perplexity = 30)
cluster.1.tsne.dt <- data.table(cluster.1.tsne$Y)
cluster.1.tsne.dt$cluster <- cluster.1.kmean$cluster
ggplot(cluster.1.tsne.dt) + geom_point(aes(x=V1, y=V2, col = as.factor(cluster))) + theme_bw()
```
```{r}
cluster.2 <- pbmc.embedet[k.mean.clusters == 2,]
cluster.2.kmean <- kmeans(cluster.2, 5)
clusters.cluster.2 <- cluster.2.kmean$cluster
cluster.2.tsne <- Rtsne(cluster.2, pca = F, perplexity = 30)
cluster.2.tsne.dt <- data.table(cluster.2.tsne$Y)
cluster.2.tsne.dt$cluster <- cluster.2.kmean$cluster
ggplot(cluster.2.tsne.dt) + geom_point(aes(x=V1, y=V2, col = as.factor(cluster))) + theme_bw()
```
```{r}
cluster.3 <- pbmc.embedet[k.mean.clusters == 3,]
cluster.3.kmean <- kmeans(cluster.3, 5)
clusters.cluster.3 <- cluster.3.kmean$cluster
cluster.3.tsne <- Rtsne(cluster.3, pca = F, perplexity = 30)
cluster.3.tsne.dt <- data.table(cluster.3.tsne$Y)
cluster.3.tsne.dt$cluster <- cluster.3.kmean$cluster
ggplot(cluster.3.tsne.dt) + geom_point(aes(x=V1, y=V2, col = as.factor(cluster))) + theme_bw()
```
```{r}
# cluster.4 <- pbmc.embedet[k.mean.clusters == 4,]
# cluster.4.kmean <- kmeans(cluster.4, 5)
# clusters.cluster.4 <- cluster.4.kmean$cluster
# cluster.4.tsne <- Rtsne(cluster.4, pca = F, perplexity = 30)
# cluster.4.tsne.dt <- data.table(cluster.4.tsne$Y)
# cluster.4.tsne.dt$cluster <- cluster.4.kmean$cluster
# ggplot(cluster.4.tsne.dt) + geom_point(aes(x=V1, y=V2, col = as.factor(cluster))) + theme_bw()
```
```{r}
# cluster.5 <- pbmc.embedet[k.mean.clusters == 5,]
# cluster.5.kmean <- kmeans(cluster.5, 5)
# clusters.cluster.5 <- cluster.5.kmean$cluster
# cluster.5.tsne <- Rtsne(cluster.5, pca = F, perplexity = 30)
# cluster.5.tsne.dt <- data.table(cluster.5.tsne$Y)
# cluster.5.tsne.dt$cluster <- cluster.5.kmean$cluster
# ggplot(cluster.5.tsne.dt) + geom_point(aes(x=V1, y=V2, col = as.factor(cluster))) + theme_bw()
```
```{r}
# cluster.6 <- pbmc.embedet[k.mean.clusters == 6,]
# cluster.6.kmean <- kmeans(cluster.6, 5)
# clusters.cluster.6 <- cluster.6.kmean$cluster
# cluster.6.tsne <- Rtsne(cluster.6, pca = F, perplexity = 30)
# cluster.6.tsne.dt <- data.table(cluster.6.tsne$Y)
# cluster.6.tsne.dt$cluster <- cluster.6.kmean$cluster
# ggplot(cluster.6.tsne.dt) + geom_point(aes(x=V1, y=V2, col = as.factor(cluster))) + theme_bw()
```
```{r}
# cluster.7 <- pbmc.embedet[k.mean.clusters == 7,]
# cluster.7.kmean <- kmeans(cluster.7, 5)
# clusters.cluster.7 <- cluster.7.kmean$cluster
# cluster.7.tsne <- Rtsne(cluster.7, pca = F, perplexity = 30)
# cluster.7.tsne.dt <- data.table(cluster.7.tsne$Y)
# cluster.7.tsne.dt$cluster <- cluster.7.kmean$cluster
# ggplot(cluster.7.tsne.dt) + geom_point(aes(x=V1, y=V2, col = as.factor(cluster))) + theme_bw()
```
```{r}
# cluster.8 <- pbmc.embedet[k.mean.clusters == 8,]
# cluster.8.kmean <- kmeans(cluster.8, 5)
# clusters.cluster.8 <- cluster.8.kmean$cluster
# cluster.8.tsne <- Rtsne(cluster.8, pca = F, perplexity = 30)
# cluster.8.tsne.dt <- data.table(cluster.8.tsne$Y)
# cluster.8.tsne.dt$cluster <- cluster.8.kmean$cluster
# ggplot(cluster.8.tsne.dt) + geom_point(aes(x=V1, y=V2, col = as.factor(cluster))) + theme_bw()
```
```{r}
# cluster.9 <- pbmc.embedet[k.mean.clusters == 9,]
# cluster.9.kmean <- kmeans(cluster.9, 5)
# clusters.cluster.9 <- cluster.9.kmean$cluster
# cluster.9.tsne <- Rtsne(cluster.9, pca = F, perplexity = 30)
# cluster.9.tsne.dt <- data.table(cluster.9.tsne$Y)
# cluster.9.tsne.dt$cluster <- cluster.9.kmean$cluster
# ggplot(cluster.9.tsne.dt) + geom_point(aes(x=V1, y=V2, col = as.factor(cluster))) + theme_bw()
```
```{r}
# cluster.10 <- pbmc.embedet[k.mean.clusters == 10,]
# cluster.10.kmean <- kmeans(cluster.10, 5)
# clusters.cluster.10 <- cluster.10.kmean$cluster
# cluster.10.tsne <- Rtsne(cluster.10, pca = F, perplexity = 30)
# cluster.10.tsne.dt <- data.table(cluster.10.tsne$Y)
# cluster.10.tsne.dt$cluster <- cluster.10.kmean$cluster
# ggplot(cluster.10.tsne.dt) + geom_point(aes(x=V1, y=V2, col = as.factor(cluster))) + theme_bw()
```
